Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-01 ind_files/99_Appendix_D_examples.tex
Significantly shorten and smoothen intro

\section{An example for quantitative traits: The quantitative logistic equation}
Recall the birth and death functionals given by \eqref{Rogers_logistic_BD}. That is, the functionals
\begin{equation}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
corresponding to an asexual population having a constant (per-capita) birth rate $r$ and mutations controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$). In terms of the scaled variable $\phi = K\nu$, these functions read:
\begin{equation}
\label{Rogers_logistic_BD_scaled}
\begin{aligned}
b_K(x|\phi) &= \frac{1}{K}b(x|\nu) = \frac{1}{K}\left( r\int\limits_{\mathcal{T}}m(x,y)K\phi(y)dy\right)\\
    d_K(x|\phi) &= \frac{1}{K}d(x|\nu) =  \frac{1}{K}\left(\frac{K\phi(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)K\phi(y)dy\right)
\end{aligned}
\end{equation}
Thus, using equation \eqref{deterministic_traj}, the deterministic trajectory becomes:
\begin{equation}
\label{Rogers_logistic_BD_deterministic}
\frac{\partial \psi}{\partial t}(x,t) = r\int\limits_{\mathcal{T}}m(x,y)\psi(y,t)dy-\frac{1}{n(x)}\psi(x,t)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y,t)dy
\end{equation}
Note that if we employ the change of variables $\Psi = K\psi$ to go back from $\mathcal{M}_{K}$ (\textit{i.e} $\phi^{(t)}$) to $\mathcal{M}$ (\textit{i.e} $\nu^{(t)}$), we recover the familiar continuous logistic equation (Equation \eqref{cts_logistic}) as the deterministic limit:
\begin{align*}
\frac{\partial \Psi}{\partial t}(x,t) &= r\int\limits_{\mathcal{T}}m(x,y)\Psi(y,t)dy-\frac{\Psi(x,t)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy \\
&\approx r\Psi(x,t) -\frac{\Psi(x,t)}{K(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy + D_m\nabla^2_{x}\Psi(x,t)
\end{align*}
where $K(x) = Kn(x)$ is the carrying capacity experienced by an individual of phenotype $x$, and $D_m = r \sigma_m^2/2$ measures the `diffusion rate' of the population in trait space. It is left as an exercise for the reader to verify by the same steps that if we instead have the birth rate functional $b(x|\phi) = \lambda\int m(x,y)\phi(y)dy$ (with $m(x,y)$ as defined in \eqref{Rogers_logistic_BD}) and the death rate functional $d(x|\phi) = \phi(x)\left(\mu+(\lambda-\mu)\phi(x)/K\right)$, the macroscopic limit yields the famous Fisher-KPP equation with growth rate $r=\lambda-\mu$ and diffusion constant $D = \lambda \sigma_m^2/2$.\\
\\
In any case, for the system defined by \eqref{Rogers_logistic_BD_scaled}, we can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}

>>>>

\section{An example for quantitative traits: The quantitative logistic equation}
Recall the birth and death functionals given by \eqref{Rogers_logistic_BD}. That is, the functionals
\begin{equation}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
corresponding to an asexual population having a constant (per-capita) birth rate $r$ and mutations controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$). In terms of the scaled variable $\phi = K\nu$, these functions read:
\begin{equation}
\label{Rogers_logistic_BD_scaled}
\begin{aligned}
b_K(x|\phi) &= \frac{1}{K}b(x|\nu) = \frac{1}{K}\left( r\int\limits_{\mathcal{T}}m(x,y)K\phi(y)dy\right)\\
    d_K(x|\phi) &= \frac{1}{K}d(x|\nu) =  \frac{1}{K}\left(\frac{K\phi(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)K\phi(y)dy\right)
\end{aligned}
\end{equation}
Thus, using equation \eqref{deterministic_traj}, the deterministic trajectory becomes:
\begin{equation}
\label{Rogers_logistic_BD_deterministic}
\frac{\partial \psi}{\partial t}(x,t) = r\int\limits_{\mathcal{T}}m(x,y)\psi(y,t)dy-\frac{1}{n(x)}\psi(x,t)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y,t)dy
\end{equation}
Note that if we employ the change of variables $\Psi = K\psi$ to go back from $\mathcal{M}_{K}$ (\textit{i.e} $\phi^{(t)}$) to $\mathcal{M}$ (\textit{i.e} $\nu^{(t)}$), we recover the familiar quantitative logistic equation as the deterministic limit:
\begin{align*}
\frac{\partial \Psi}{\partial t}(x,t) &= r\int\limits_{\mathcal{T}}m(x,y)\Psi(y,t)dy-\frac{\Psi(x,t)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy \\
&\approx r\Psi(x,t) -\frac{\Psi(x,t)}{K(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy + D_m\nabla^2_{x}\Psi(x,t)
\end{align*}
where $K(x) = Kn(x)$ is the carrying capacity experienced by an individual of phenotype $x$, and $D_m = r \sigma_m^2/2$ measures the `diffusion rate' of the population in trait space. It is left as an exercise for the reader to verify by the same steps that if we instead have the birth rate functional $b(x|\phi) = \lambda\int m(x,y)\phi(y)dy$ (with $m(x,y)$ as defined in \eqref{Rogers_logistic_BD}) and the death rate functional $d(x|\phi) = \phi(x)\left(\mu+(\lambda-\mu)\phi(x)/K\right)$, the macroscopic limit yields the famous Fisher-KPP equation with growth rate $r=\lambda-\mu$ and diffusion constant $D = \lambda \sigma_m^2/2$.\\
\\
In any case, for the system defined by \eqref{Rogers_logistic_BD_scaled}, we can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}
